Closing the empirical cycle

### 2022-11-10 manuscript/manuscript.Rmd
update 2022-11-10 11:47:07

abstract: |
  | **Research Highlights**
  | • Individual differences (e.g., in personality, motivation, and empathy) are among the most important predictors of emotion regulation development.
  | • Specific parenting practices may be less relevant than previously thought, and negative parenting practices may be more impactful than positive ones.
  | • Important predictors most suitable for early risk assessment include regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior.
  | • Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
  | [As 20% of adolescents develop emotion regulation difficulties, it is important to identify important early predictors thereof. Using the machine learning algorithm SEM-forests, we ranked the importance of (87) candidate variables assessed at age 13 in predicting quadratic latent trajectory models of emotion regulation development from age 14-18. Participants were 497 Dutch adolescents. Results indicated that the most important predictors were individual differences (e.g., in personality), aspects of relationship quality and conflict behaviors with parents and peers, and internalizing and externalizing problems. Relatively less important were demographics, bullying, delinquency, substance use, and specific parenting practices – and negative parenting practices ranked higher than positive ones. We discuss implications for theory and interventions, and present an open source risk assessment tool, ERRATA.]{#abstract}

@@

A key challenge in adolescence is developing mature emotion regulation abilities [@zimmermannEmotionRegulationEarly2014]. 
Adolescents experience biological, cognitive, and social changes that create an opportunity for positive emotion regulation development [@steinbergAgeOpportunityLessons2014].
Still, [as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
with potentially lifelong implications for mental health and well-being [@aldaoEmotionregulationStrategiesPsychopathology2010].
These difficulties in emotion regulation can cause substantial distress to the individual,
and incur a sizable societal cost via their implications for mental health [@leeAdolescentMentalHealth2014].]{#cost}
A comprehensive understanding of the risk factors that predispose some adolescents to emotion dysregulation could aid in the development of effective screening and interventions.
<span id = "abundantlit">There is already an abundance of empirical research on adolescent emotion regulation and its predictors; a recent systematic review identified 6305 relevant publications [@vanlissaMappingPhenomenaRelevant2021].
Still, that review identified specific limitations of the existing literature:
many studies narrowly focus on a few predictors within a specific research area,
and thus fail to provide insight into these predictors' relative importance compared to other predictors, including those in other domains.
Relatedly,
the literature lacks a unifying theoretical framework that is both specific to adolescence and comprehensive in scope [see @bariolaChildAdolescentEmotion2011].</span>

The present study seeks to overcome these limitations
by using data-driven insights to complement and reflect on existing theory.
To this end, a broad range of potential risk factors were analyzed using machine learning,
and ranked in terms of predictive importance.
Specifically,
<span id = "personcentered1">the SEM-forests algorithm [@brandmaierTheoryguidedExplorationStructural2016], a person-centered analysis technique, was used
to identify the most important early indicators (at age 13) of developmental trajectories
of emotion regulation across adolescence (age 14 through 18).
The results indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle through adolescence [@croneUnderstandingAdolescencePeriod2012].</span>
These insights can guide development of interventions that focus on the most important predictors,
and direct the formation of overarching theory that adequately represents essential phenomena.

@@

Longitudinal studies of emotion regulation development in adolescence provide preliminary empirical support for this notion of diverging trajectories: 
Such studies consistently reveal large between-individual heterogeneity [see @craccoEmotionRegulationChildhood2017; @vanlissaRoleFathersMothers2019; @zimmermannEmotionRegulationEarly2014; @reindlSocializationEmotionRegulation2016].
<!-- Although some studies have examined, on average, how strongly particular risk factors are associated with developmental trajectories - most have done  REF write something about other studies examining risk factors -->
An important next step is to identify which risk factors render some adolescents susceptible to emotion dysregulation.
<!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards "person-centered" research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23. Seminal studies have applied "clustering" methods to identify youth with similar developmental trajectories14,24–26. Adolescents may follow similar trajectories for different reasons, however, and person-centered methods are too complex to accommodate more than a handful of theoretically-derived predictors. Unfortunately, relevant theory on adolescent emotion regulation development is limited27–29. The field is therefore at an impasse: We know that some adolescents are more susceptible to emotion dysregulation than others, but lack tools and theory to identify important predictors of individual development30. -->

@@

A recent review noted that the literature on potential predictors of emotion regulation development is somewhat fragmented,
with most studies considering only a handful of variables within the scope of a particular sub-discipline [@vanlissaMappingPhenomenaRelevant2021].
<!-- At the same time, this literature is relatively fragmented, -->
<!-- as most studies have examined only a limited number of predictors [@vanlissaMappingPhenomenaRelevant2021]. -->
This fragmentation may be due, in part, to the fact that few theories are tailored specifically to adolescence,
and none consistently guide research in the field [@bussTheoriesEmotionalDevelopment2019].
Theories commonly cited in this field vary in scope and specificity:
Some are broad in scope, considering many potential risk factors,
but are somewhat non-specific.
This limits their usefulness in generating testable hypotheses,
and renders them impervious to falsification or improvement.
Other theories are narrower in scope, and are sufficiently specific to derive testable predictions, but lack broader context.
For a more detailed overview, see @vanlissaMappingPhenomenaRelevant2021.
Two of the most influential broad theories are the bioecological model [@bronfenbrennerBioecologicalModelHuman2007; e.g., see @augustEmotionRegulationAtRisk2017] and the transactional model [@sameroffUnifiedTheoryDevelopment2010].
The bioecological model describes how individual development is shaped by factors at different levels of analysis:
The individual level, which includes biological make-up and individual differences;
the microsystem, composed of primary caregivers and close relationships;
the mesosystem, including school and neighborhood factors;
the exosystem, including media, political- and economic influences;
and the macrosystem, consisting of cultural norms and values.
The transactional model is compatible with the bioecological model,
and describes development as an interplay between the individual child and the environment.
It distinguishes proximal environmental influences, roughly corresponding to the microsystem,
from more distal influences.
With increasing age, distal influences are thought to become more influential.

More specific theories tend to focus on cognitive maturation and the role of primary caregivers.
Examples include the theory of normative emotional development [@sroufeAttachmentDyadicRegulation1996]
and the tripartite model of emotion regulation socialization in families [@morrisRoleFamilyContext2007], which has also been applied to the peer context [@reindlSocializationEmotionRegulation2016].
This tripartite model is one of the few theories that explicitly addresses the life phase of adolescence.<!---->
<!-- This seems a bit selective. For example, also discuss the work of Hollenstein and of Nicholas Allen. -->
<!-- I think you do insufficient right to the theories that exist, both in breadth and in depth. Perhaps just stay closer to the contribution of the current paper: empirically examine multiple factors simultaneously rather than selectively examining a few? -->
<!-- Despite this abundance of relevant theory,  -->
<!-- several reviews have noted the relatively fragmented nature of the literature  -->
<!-- that  -->
<!-- few theories are tailored specifically to emotion regulation development in adolescence, -->
<!-- and none are used consistently to guide research in this field. -->
Given that few theories are both comprehensive and sufficiently specific to derive testable hypotheses,
an important challenge for the field is to work towards an integrating theory of emotion regulation development in adolescence.

<!-- * Bioecological model [@bronfenbrennerBioecologicalModelHuman2007] -->
<!-- * Transactional model [@sameroffUnifiedTheoryDevelopment2010] -->
<!-- * Hall's notion of "storm and stress" -->
<!-- * Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995] -->
<!-- * Tripartite model [@morrisRoleFamilyContext2007] -->
<!-- * Internalization model [@holodynskiDevelopmentEmotionsEmotion2006] -->
<!-- * Polyvagal theory [@porgesOrientingDefensiveWorld1995]. -->
<!-- * Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012] -->
<!-- * Process model of emotion regulation [@grossHandbookEmotionRegulation2013] -->
<!-- * Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000] -->
<!-- Maybe review theory briefly? -->
The first step towards theory formation is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
<!--Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
A recent literature review did so
using a text mining systematic review of author keywords and abstracts of 6305 publications on  adolescents’ emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
Typically, phenomena are identified using narrative literature reviews.
However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
Specifically, related keywords are grouped into clusters,
with each cluster representing a phenomenon.
These phenomena are then ranked by the number of studies those keywords occur in, and potential relations between phenomena are mapped by co-occurrence within studies.<!-- reflection and meaning-making-->
This review thus offers a relatively comprehensive overview of potentially relevant phenomena.
The present study used it as a starting point for an empirical investigation of the relative importance and direction of effect of these phenomena.

According to this prior review,
the phenomena that have received most research attention are internalizing and externalizing problems, and mental health.
These are likely to be, at least in part, outcomes of emotion dysregulation.
<!-- - but they that does not diminish their usefulness as predictors. -->
Other important phenomena reflected what has been discussed in prior narrative reviews,
including cognitive development, parenting, and external stressors [cf. @coe-odessEmergentEmotionsAdolescence2019].
Finally, the method identified some phenomena that have received less attention in prior theory and reviews,
such as substance use, structural disadvantage,
identity, morality, sex and sexual development.
<!-- To this end, the present study sought to identify operationalizations of these relevant phenomena in a large-scale panel study. -->
Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development.

What is still missing is an understanding of the relative utility of different early indicators in predicting adolescents' developmental outcomes.
<!-- we do not know the relative importance of  with respect to others not included in the same study. -->
Put simply, each study has examined a piece of the puzzle,
but no single study has captured the complete picture.
This knowledge gap cannot be addressed by comparing the relative importance of different predictors across studies,
as different studies tend to sample from different populations,
use different research designs,
different control variables and statistical models.
Furthermore, conventional methods are ill-suited to overcome address this research question:
<!-- In part, this limitation is a consequence of conventional methods: -->
linear models limit the total number of predictors that can be included due to concerns of power, overfitting, and multicolinearity [@hastieElementsStatisticalLearning2009].
By contrast, some machine learning methods offer a more holistic approach,
as they can incorporate many candidate predictors and perform variable selection,
while incorporating checks and balances to ensure generalizable results [@hastieElementsStatisticalLearning2009].
The present study used a specific machine learning algorithm uniquely suited for theory-guided exploration: SEM-forests [@brandmaierTheoryguidedExplorationStructural2016].

@@

<span id = "semforests">The SEM-forests algorithm is based on the random forests algorithm, which in turn is based on regression trees [see @stroblIntroductionRecursivePartitioning2009].
A single regression tree splits the sample repeatedly in two by whichever predictor $p$ makes both post-split groups maximally internally homogenous, and maximally different from one another.
A random forest consists of many regression trees grown on bootstrapped samples from the original dataset.
To ensure that each tree learns something different from the data, it may only consider a random subsample of $K$ candidates out of all $p$ predictors at each split.
With this in mind, the SEM-forest algorithm operates as follows:

1. Bootstrap the data with replacement; we used 1000 bootstrap replications
2. For each bootstrap sample, construct a SEM tree:
    i. Estimate the latent growth model on the bootstrap sample
    ii. Consider a random selection of $K$ candidates out of all $p$ predictors; we used $k = 9$ based on the standard practice to use $K = \lfloor\sqrt{p}\rfloor$
    iii. [Identify the predictor $k$ and value of that predictor that, when using that predictor and value to split the sample in two, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).]{#stepiii}
    iv. [Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then fully grown.]{#fullygrown}
3. [Given an individual's scores on the predictor variables $p$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters). Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.]{#predictions}

Each tree thus maps relationships between the predictors and outcome as a series of recursive splits of the data.
This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
<span id = "examplenonlinear">As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.</span>
The algorithm additionally performs variable selection;
unimportant variables are not selected in step 2.iii.
Thus, the number of predictors can be large without risking model non-identification.</span>

@@

<!-- SB: you have convincingly argued why machine learning in general is important, yet I miss the details related to ER. I would recommend to add a paragraph summarizing theories of ER and the factors they consider important. Also, describe the candidate predictors that were derived from the literature search, and how they relate to theories and /or empirical evidence -->
The present study set out to obtain a relatively comprehensive understanding of the predictors of emotion regulation development across adolescence.
<!-- identify the most important early indicators of developmental trajectories -->
<!-- of  -->
The primary aim was to identify the most important predictors of adolescents' developmental trajectories of emotion regulation development.
The secondary aim was to probe the nature of the relationship of each predictor with these trajectories.
This involves an important challenge, as
the number of potentially relevant risk factors is too large for conventional methods.
At the same time, there is a paucity of relevant theory to narrow down the candidate predictors a priori.
<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
This challenge can be overcome by machine learning methods that perform variable selection.
We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.

@@

`r ifelse(mask, "The Workflow for Open Reproducible Code in Science (WORCS)", "**Masked**")` was used to make analyses reproducible  `r ifelse(mask, "[@vanlissaWORCSWorkflowOpen2020]", "**(masked reference)**")`.
All analysis code, Supplemental materials and synthetic data are available at `r ifelse(mask, "<https://github.com/cjvanlissa/veni_forest>", "**masked URL**")`.
The original data (controlled access) and all study documentation are available via `r ifelse(mask, "<https://doi.org/10.17026/DANS-ZRB-V5WP>", "**masked URL**")`.

@@

```{r}
desc_num <- read.csv("../desc_num.csv", stringsAsFactors = FALSE)
desc_cat <- read.csv("../desc_cat.csv", stringsAsFactors = FALSE)
```
<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which 
received ethical approval from the ethical committee of the University Medical Centre `r ifelse(mask, "Utrecht", "**masked**")`.
The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.</span>
Of 1,544 eligible families, 497 produced informed consent and were included in the study.
[From 2006 to 2011, trained interviewers conducted six annual measurement waves to collect questionnaire data during home visits.]{#waves}
All independent variables were assessed in the first measurement wave.
[The sample demographic information is reported in Tables \@ref(tab:tabcat) and \@ref(tab:tabnum).]{#demographics}
The dependent variable was measured from the second to the sixth wave.
Families received €100 financial compensation for their participation in annual measurements.

<!-- **Participants** -->

<!-- * RADAR data (https://www.uu.nl/en/research/radar) -->
<!-- * 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) -->
<!-- * Collected between 2006-2011 -->
<!-- * Most families were  medium- to high-SES (10% low-SES) -->

@@

## Dependent variable

@@

From wave 2-6, adolescents completed the Difficulties in Emotion Regulation Scale, which was reverse coded to represent emotion regulation [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2019],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
[To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.]{#singlefactor}
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the sample average trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
[Note that the RMSEA has been found to be more appropriate for longitudinal data, possibly because the null-model on which the CFI is based is less appropriate for repeated measures [@leiteDetectingGrowthShape2011]]{#cfi}.
As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we considered it acceptable to proceed with individual factor scores for further analysis.

## Candidate predictors

>>>>

abstract: |
  | **Research Highlights**
  | • Individual differences (e.g., in personality, motivation, and empathy) are among the most important predictors of emotion regulation development.
  | • Specific parenting practices may be less relevant than previously thought, and negative parenting practices may be more impactful than positive ones.
  | • Important predictors most suitable for early risk assessment include regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior.
  | • Future theory should emphasize proximal- over distal processes, focus on good-enough parenting, and reassess common perceptions about “risky behaviors” like delinquency, substance use, and bullying.
  | [As 20% of adolescents develop emotion regulation difficulties, it is important to identify important early predictors thereof. Using the machine learning algorithm SEM-forests, we ranked the importance of (87) candidate variables assessed at age 13 in predicting quadratic latent trajectory models of emotion regulation development from age 14-18. Participants were 497 Dutch families. Results indicated that the most important predictors were individual differences (e.g., in personality), aspects of relationship quality and conflict behaviors with parents and peers, and internalizing and externalizing problems. Relatively less important were demographics, bullying, delinquency, substance use, and specific parenting practices – although negative parenting practices ranked higher than positive ones. We discuss implications for theory and interventions, and present an open source risk assessment tool, ERRATA.]{#abstract}

@@

A key challenge in adolescence is developing mature emotion regulation abilities [@zimmermannEmotionRegulationEarly2014].
Adolescents experience biological, cognitive, and social changes that create an opportunity for positive emotion regulation development [@steinbergAgeOpportunityLessons2014].
Still, [as many as 20% of adolescents instead experience severe emotion regulation difficulties [@leeAdolescentMentalHealth2014],
with potentially lifelong implications for mental health and well-being [@aldaoEmotionregulationStrategiesPsychopathology2010].
These difficulties in emotion regulation can cause substantial distress to the individual,
and incur a sizable societal cost via their implications for mental health [@leeAdolescentMentalHealth2014].]{#cost}
A comprehensive understanding of the risk factors that predispose some adolescents to emotion dysregulation could aid in the development of effective screening and interventions.
<span id = "abundantlit">There is already an abundance of empirical research on adolescent emotion regulation and its predictors; a recent systematic review identified 6305 relevant publications [@vanlissaMappingPhenomenaRelevant2021].
An important limitation of the existing literature is that many studies narrowly focus on a few predictors within a specific research area,
and thus fail to provide insight into these predictors' relative importance compared to other predictors, including those in other domains.
Relatedly,
the literature lacks a unifying theoretical framework that is both specific to adolescence and comprehensive in scope [see @bariolaChildAdolescentEmotion2011].</span>

This exploratory study seeks to overcome these limitations by casting a broad net among potential risk factors, 
and using the resulting data-driven insights to complement and reflect on existing theory.
Machine learning was used to determine the relative utility of a wide range of potential risk factors.
Specifically,
<span id = "personcentered1">the SEM-forests algorithm [@brandmaierTheoryguidedExplorationStructural2016], a person-centered analysis technique, was used
to identify the most important early indicators (at age 13) of developmental trajectories
of emotion regulation across adolescence (age 14 through 18).
The results indicate which factors are most predictive of adolescents' diverging trajectories of emotion regulation development, contributing to our understanding of why some youth flourish while others struggle [@croneUnderstandingAdolescencePeriod2012].</span>
These insights can guide development of interventions that focus on the most important predictors,
and direct the formation of overarching theory that adequately represents essential phenomena.

@@

Longitudinal studies of emotion regulation development in adolescence provide preliminary empirical support for this notion of diverging trajectories: 
Such studies consistently reveal large between-individual heterogeneity [see @craccoEmotionRegulationChildhood2017; @vanlissaRoleFathersMothers2019; @zimmermannEmotionRegulationEarly2014; @reindlSocializationEmotionRegulation2016].
<!-- Although some studies have examined, on average, how strongly particular risk factors are associated with developmental trajectories - most have done  REF write something about other studies examining risk factors -->
[An important target of research is thus to identify which risk factors render some adolescents susceptible to emotion dysregulation.]{#nextstep}
<!-- This goal fits within the paradigm shift that has gripped the field of pedagogy, towards "person-centered" research, which focuses on individual developmental trajectories and within-family predictive processes, rather than average patterns across samples16,23. Seminal studies have applied "clustering" methods to identify youth with similar developmental trajectories14,24–26. Adolescents may follow similar trajectories for different reasons, however, and person-centered methods are too complex to accommodate more than a handful of theoretically-derived predictors. Unfortunately, relevant theory on adolescent emotion regulation development is limited27–29. The field is therefore at an impasse: We know that some adolescents are more susceptible to emotion dysregulation than others, but lack tools and theory to identify important predictors of individual development30. -->

@@

[Substantial prior research has examined potential risk factors.
However, a recent review of 6305 relevant publications noted that this literature is somewhat fragmented,
with most studies considering only a handful of variables within the scope of a particular sub-discipline [@vanlissaMappingPhenomenaRelevant2021].
If each study examines only a piece of the puzzle, it is hard to see the big picture.]{#fragmented}
This fragmentation may be due, in part,
to the fact that few theories are tailored to emotion regulation in adolescence,
and none consistently guide research in the field [@bussTheoriesEmotionalDevelopment2019].
Commonly cited theories in this field vary in scope and specificity [see @vanlissaMappingPhenomenaRelevant2021]:
Some are broad in scope, considering many potential risk factors,
but are somewhat non-specific.
This limits their usefulness in generating testable hypotheses,
and renders them impervious to falsification or improvement.
Other theories are narrower in scope, and are sufficiently specific to derive testable predictions, but lack broader context.
Two of the most influential broad theories are the bioecological model [@bronfenbrennerBioecologicalModelHuman2007; e.g., see @augustEmotionRegulationAtRisk2017] and the transactional model [@sameroffUnifiedTheoryDevelopment2010].
The bioecological model describes how individual development is shaped by factors at different levels of analysis:
The individual level, which includes biological make-up and individual differences;
the microsystem, composed of primary caregivers and close relationships;
the mesosystem, including school and neighborhood factors;
the exosystem, including media, political- and economic influences;
and the macrosystem, consisting of cultural norms and values.
<span id = "velez1">Note that a recent perspective on the bioecological model emphasizes that cultural practices are enacted by individuals at different levels of the model, including parents [@velez-agostoBronfenbrennerBioecologicalTheory2017].
This perspective is consistent with the transactional model,
which describes development as an interplay between the individual child and the environment.</span>
It distinguishes proximal environmental influences, roughly corresponding to the microsystem,
from more distal influences.
With increasing age, distal influences are thought to become more influential.

More specific theories tend to focus on cognitive maturation and the role of primary caregivers.
Examples include the theory of normative emotional development [@sroufeAttachmentDyadicRegulation1996]
and the tripartite model of emotion regulation socialization in families [@morrisRoleFamilyContext2007], which has also been applied to the peer context [@reindlSocializationEmotionRegulation2016].
The tripartite model is one of the few theories that explicitly addresses the life phase of adolescence.
It posits that children's emotion regulation is shaped by modeling, specific parenting practices, and the emotional family climate (including the interparental relationship).
This model is relatively narrow, focusing on parents' role - but also sufficiently specific to derive testable hypotheses.
Note that few theories are both comprehensive and specific.
An important challenge for the field is thus to work towards an integrating theory of emotion regulation development in adolescence.

The first step towards theory formation is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
Phenomena considered relevant in a field of research are often summarized using literature reviews.
A recent review of 6305 publications on adolescents’ emotion regulation used text mining to identify relevant phenomena, including potential risk factors [@vanlissaMappingPhenomenaRelevant2021].
Narrative reviews are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
Text mining systematic reviews are more comprehensive and derive insights from the literature through a transparent and somewhat objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021].
This review thus offers a relatively comprehensive overview of potential predictors of emotion regulation development.
The present study used it as a starting point for an empirical investigation of the relative importance and direction of effect of these predictors.

According to this prior review,
the phenomena that have received most research attention are internalizing and externalizing problems, and mental health.
Although reciprocal effects cannot be excluded, these phenomena are typically conceptualized as outcomes of emotion dysregulation.
Note, however, that this does not diminish their usefulness as predictors.
Other important phenomena reflected what has been discussed in prior narrative reviews,
including cognitive development, parenting, and external stressors [cf. @coe-odessEmergentEmotionsAdolescence2019].
Finally, the method identified some phenomena that have received less attention in prior theory and reviews,
such as substance use, structural disadvantage,
identity, morality, sex and sexual development.
Prior theory and empirical research thus suggests many phenomena potentially related to heterogeneity in adolescents' emotion regulation development.

The aforementioned potential risk factors have largely been studied in isolation.
What is still missing is an understanding of the relative utility of different early indicators in predicting adolescents' developmental outcomes.
This knowledge gap cannot be addressed by comparing the relative importance of different predictors across studies,
as different studies tend to sample from different populations,
use different research designs,
different control variables and statistical models.
<!-- Furthermore, conventional methods are ill-suited to address this research question: -->
In part, this knowledge gap is a consequence of the limitations of conventional methods:
linear models limit the total number of predictors that can be included due to power concerns, risk of overfitting, and multicolinearity [@hastieElementsStatisticalLearning2009].
By contrast, some machine learning methods offer a more holistic approach,
as they can incorporate many candidate predictors and perform variable selection,
while incorporating checks and balances to ensure generalizable results [@hastieElementsStatisticalLearning2009].
The present study used a specific machine learning algorithm uniquely suited for theory-guided exploration: SEM-forests [@brandmaierTheoryguidedExplorationStructural2016].

@@

<span id = "semforests">The SEM-forests algorithm is based on the random forests algorithm, which in turn is based on regression trees [for an accessible introduction, see @stroblIntroductionRecursivePartitioning2009].
A single regression tree splits the sample repeatedly in two by whichever predictor makes both post-split groups maximally internally homogenous, and maximally different from one another.
A random forest consists of many regression trees grown on bootstrapped samples from the original dataset.
To ensure that each tree learns something different from the data, it may only consider a random subsample of $m$ candidates out of all $p$ predictors at each split.
With this in mind, the SEM-forest algorithm operates as follows:

1. Bootstrap the data with replacement; we used 1000 bootstrap replications
2. For each bootstrap sample, construct a SEM tree:
    i. Estimate the latent growth model on the bootstrap sample
    ii. Consider a random selection of $m$ candidates out of all $p$ predictors; commonly $m = \lfloor\sqrt{p}\rfloor$
    iii. [Identify the predictor $p_k$ and value of that predictor $x$ that, when using individual values on that predictor $p_{ki}$ to split the sample in two by the rule $p_{ki} \leq x$, maximizes the difference in model parameters between the two post-split groups (as determined by a likelihood ratio test).]{#stepiii}
    iv. [Build a SEM-tree by repeating steps 3 and 4 until all post-split groups reach a minimum size that is too small for further splitting. The tree is then considered fully grown.]{#fullygrown}
3. [Given an individual's scores on all predictor variables $p_i$, we can now traverse each tree in the forest to obtain a predicted value of the model parameters (in this case, the latent growth parameters). Aggregating the predicted parameters across all SEM-trees in the forest averages out prediction error.]{#predictions}

Each tree thus maps relationships between the predictors and outcome as a series of recursive splits of the data.
This approach intrinsically accommodates non-linear effects and (complex) interactions, which is an advantage over parametric models that assume linearity.
[As a simple example, imagine that only very low levels of autonomy support would be associated with elevated levels of emotion regulation (i.e., a higher intercept).
In regression analysis, this could be represented with a linear and quadratic term, and a plot might show that the effect of autonomy support on the intercept increases to a value $x_{knee}$, and then tapers off.
A tree-based model would represent the same effect by splitting the sample on autonomy support at the value $x_{knee}$.]{#nonlinearexample}
Note that SEM-forests intrinsically perform variable selection: unimportant variables are not selected in step 2.iii.
Thus, the number of predictors can be large without risking model non-identification.</span>

@@

The present study set out to obtain a comprehensive understanding of the relative utility of predictors of emotion regulation development across adolescence.
The primary aim was to identify the most important predictors of adolescents' developmental trajectories of emotion regulation development.
The secondary aim was to probe the nature of the relationship of each predictor with these trajectories.
An important challenge of these research aims is that the number of potentially relevant risk factors is too large for conventional methods.
At the same time, there is a paucity of relevant theory to narrow down the candidate predictors a priori.
We sought to overcome this challenge by using SEM-forests, 
a machine learning algorithm methods that performs variable selection,
to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence.

@@

`r ifelse(mask, "The Workflow for Open Reproducible Code in Science (WORCS)", "**Masked**")` was used to create a reproducible research archive  `r ifelse(mask, "[@vanlissaWORCSWorkflowOpen2020]", "**(masked reference)**")`.
All analysis code, supplemental materials and synthetic data are available at `r ifelse(mask, "<https://github.com/cjvanlissa/veni_forest>", "**masked URL**")`.
The original data and all study documentation are available under controlled access via `r ifelse(mask, "<https://doi.org/10.17026/DANS-ZRB-V5WP>", "**masked URL**")`.

@@

```{r}
desc_num <- read.csv("../desc_num.csv", stringsAsFactors = FALSE)
desc_cat <- read.csv("../desc_cat.csv", stringsAsFactors = FALSE)
```
<span id = "dutch">Participants were 497 Dutch adolescents and their parents enrolled in the ongoing longitudinal study `r ifelse(mask, "RADAR [@branjeResearchAdolescentDevelopment2018]", "**(masked reference)**")`, which 
received ethical approval from the ethical committee of the University Medical Centre `r ifelse(mask, "Utrecht", "**masked**")`.
The `r ifelse(mask, "RADAR", "**masked**")` sample was recruited from over 200 randomly selected elementary schools in the province of `r ifelse(mask, "Utrecht", "**masked**")`, and four main cities in `r ifelse(mask, "The Netherlands", "**masked**")`.</span>
Of 1,544 eligible families, 497 produced informed consent and were included in the study.
[From 2006 to 2011, trained interviewers completed six annual measurement waves by collecting questionnaire data during home visits.]{#waves}
All independent variables were assessed in the first measurement wave.
[Sample demographic characteristics are reported in Tables \@ref(tab:tabcat) and \@ref(tab:tabnum).]{#demographics}
The dependent variable was measured from the second to the sixth wave.
Families received €100 financial compensation for their participation in annual measurements.

@@

### Dependent variable

@@

From wave 2-6, adolescents completed the Difficulties in Emotion Regulation Scale, which was reverse coded to represent emotion regulation [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2019],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
[To this end, we estimated a CFA with all DERS items loading on one latent variable per measurement wave.]{#singlefactor}
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the sample average trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the metric invariance model indicated acceptable fit ($\text{RMSEA} = .05$,  $\text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
[Note that the RMSEA has been found to be more appropriate for longitudinal data, possibly because the null-model on which the CFI is based is less appropriate for repeated measures [@leiteDetectingGrowthShape2011]]{#cfi}.
As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we proceeded with individual factor scores for further analysis.

### Candidate predictors
